Accelerating adaptive ultrasound algorithms using graphics processing units \vspace{0.5cm}

### 2014-02-12 ThesisJP/Background/background.tex
updated cpu vs gpu figure.
Some writing on the background.

Around the same time as CPU manufactures hit the frequency wall, people started experimenting with programmable shaders which recently had been introduced in the field of graphics programming. The graphics processing pipeline had before this been a fixed function pipeline, where e.g. geometry and textures where fed to the graphics processing unit (GPU) in one end and a rasterized rendered image was outputted in the other. What programmable shaders  introduced was the ability to replace some steps in the fixed-function pipeline with custom computations. This made it possible to utilize the GPU for other tasks than just rendering graphics \cite{Seland2007}. The rationale behind this exploration was that GPUs were already a highly available processor, and hidden behind its graphics interface was processing powers an order of magnitude larger than what the CPU could provide (Fig.\,\ref{fig:cpu_vs_gpu}). Actually, at the time when CPUs hit the frequency wall the GPU had already gone multi-core, driven by an ever-increasing demand for more realistic computer games. The rendering pipeline exhibits minimal execution path branching, and data are typically used once. Designers of GPUs could therefore skip advance feature, found in most CPUs, like branch prediction and different levels of data caching. This reduced the silicon footprint of each core and made it possible to add multiple cores to the same chip before CPU designers where able to do the same. Today, even if more advanced caching has been added to GPUs, this is still the main reason why GPUs have higher peak performance than CPUs (Fig.\,\ref{fig:cpu_vs_gpu}). Another reason is the inherent parallel nature of the rendering problem. When geometry is projected and rasterized the same instruction is typically needed for a lot of data at the same time. This makes a special kind of instruction, Single Instruction Multiple Data (\nom{SIMD}{Single instruction multiple data (instruction)}) instruction, especially suited for this kind of problem. The differences between how SIMD instructions are utilized in modern CPUs and GPUs is discussed in Section \ref{sec:cpu_vs_gpu}.

For the first adopters of GPU programming, or General-purpose programming on graphics processing units (GPGPU) as it was named, it soon became evident that offloading all types of computations to the GPU was not always a good idea. In most cases it is important to balance the load equally between the CPU and GPU to obtain their combined computationally power. Balancing typically means that highly parallel tasks should be offloaded to the GPU and serialized work should stay on the CPU. This paradigm of utilizing specialized cores for solving specific problems is today known as heterogeneous computing. The specialized core can

FPGAs (Field-programmable gate array), APU (Accelerated processing unit).

Three currently ongoing transitions: Multi-core, Heterogeneous-cores, and Cloud-cores.

This thesis is focusing on heterogeneous computing with CPUs and GPUs.

@@

@@

\begin{figure}
\centering
\includegraphics[width=0.9\textwidth]{img/cpu-vs-gpu.png}
\caption{Development in theoretical peak throughput, single precision (SP) and double precision (DP), the last decade for Intel CPUs and Nvidia GPUs. Image courtesy of Michael Galloy (michaelgalloy.com).}
\label{fig:cpu_vs_gpu}
\end{figure}

@@

>>>>

Around the same time as CPU manufactures hit the frequency wall, people started experimenting with programmable shaders which recently had been introduced in the field of graphics programming. The graphics processing pipeline had before this been a fixed function pipeline, where e.g. geometry and textures where fed to the graphics processing unit (GPU) in one end and a rasterized rendered image was outputted in the other. What programmable shaders  introduced was the ability to replace some steps in the fixed-function pipeline with custom computations. This made it possible to utilize the GPU for other tasks than just rendering graphics \cite{Seland2007}. The rationale behind this exploration was that GPUs were already highly available in consumer PCs, and hidden behind its graphics interface were processing powers an order of magnitude larger than what the CPU could provide (Fig.\,\ref{fig:cpu_vs_gpu}). Actually, at the time when CPUs hit the frequency wall the GPU had already gone multi-core, first of all driven by an ever-increasing demand for more realistic computer games. Second, the rendering pipeline exhibits minimal execution path branching, and data are typically used once. Designers of GPUs could therefore skip advance feature, found in most CPUs, like branch prediction and different levels of data caching. These measures reduced the silicon footprint of each core and made it possible to add multiple cores to one GPU before CPU designers where able to do the same. Today, even if more advanced caching has been added to GPUs, this is still the main reason why GPUs have higher peak performance than CPUs (Fig.\,\ref{fig:cpu_vs_gpu}). Another reason is the inherent parallel nature of the rendering problem. When geometry is projected and rasterized the same instruction is typically needed for a lot of data at the same time. This makes a special kind of instruction, Single Instruction Multiple Data (\nom{SIMD}{Single instruction multiple data (instruction)}) instruction, especially suited for this kind of problem. The differences between how SIMD instructions are utilized in modern CPUs and GPUs are discussed in Section \ref{sec:cpu_vs_gpu}.

For the first adopters of GPU programming, or General-purpose programming on graphics processing units (GPGPU) as it was named, it soon became evident that offloading all types of computations to the GPU was not always a good idea. In most cases it is important to balance the load equally between the CPU and GPU to obtain their combined computationally power. Balancing typically means that highly parallel tasks should be offloaded to the GPU and serialized work should stay on the CPU. This paradigm of utilizing specialized cores for solving specific problems is today known as heterogeneous computing. The specialized core can be a GPU or e.g. a field-programmable gate array (\nom{FPGA}{Field-programmable gate array}). The latest trend is that specialized accelerator cores are embedded onto the CPU. This heterogeneous processors is often referred to as an accelerated processing unit (\nom{APU}{Accelerated processing unit}). The last three generations of CPUs from Intel have all had on-chip integrated graphics.

The final transition currently taking place in the computer industry is the introduction of cloud computing, where compute power is provided as a web-service and automatically scales on demand. Together, multi-core, heterogeneous cores, and cloud cores forms the "jungle" that todays software engineers have to navigate (Fig.\,\ref{fig:jungle}).

This thesis focus on heterogeneous computing with one multi-core CPU and a discrete high performance GPU, a cheap system which is likely to be the specification of current and future ultrasound imaging system.

@@

 the most powerful one (HD 5200) have a peak performance of nearly one tera single precision floating point operations per second (FLOPs).

@@

\begin{figure}
\centering
\includegraphics[width=\textwidth]{img/cpu_vs_gpu.pdf}
\caption{Development in theoretical peak throughput, single precision (SP) and double precision (DP), the last decade for Intel CPUs and Nvidia GPUs. Image courtesy of Michael Galloy (michaelgalloy.com).}
\label{fig:cpu_vs_gpu}
\end{figure}

@@

Sampling rates, and data size.
%Due to the high bandwidth requirement for medical ultrasound imaging, as we will see in Section \ref{sec:ultrasound}, a cloud computing ultrasound imaging system will not feasibly for a long time.
Because of the high data rate and processing requirements ultrasound systems where first analog systems up to the mid 90s. The first digital systems where made with custom boards and ASICs.
